---
title: "Brain Cancer Microarray Data Weighted Gene Co-expression Network Analysis R Tutorial "
author: "Steve Horvath, Bin Zhang, Jun Dong, Tova Fuller, Peter Langfelder"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
setwd("/Users/samoz/Documents/phd/year1/stat 465/project/STAT_465_gene_coexpression/tutorial")
```

# CONTENTS

This is stolen from: ![https://horvath.genetics.ucla.edu/html/GeneralFramework/GBMTutorialHorvath.pdf]

This document contains function for carrying out the following tasks

A) Assessing scale free topology and choosing the parameters of the adjacency function
    using the scale free topology criterion (Zhang and Horvath 05)

B) Computing the topological overlap matrix

C) Defining gene modules using clustering procedures

D) Summing up modules by their first principal component (first eigengene)

E) Relating a measure of gene significance to the modules

F) Carrying out a within module analysis (computing intramodular connectivity)
    and relating intramodular connectivity to gene significance.

G) Miscellaneous other functions, e.g. for computing the cluster coefficient.


# Downloading the R software
1) Go to http://www.R-project.org, download R and install it on your computer

After installing R, you need to install several additional R library packages:

To get this tutorial and data files, go to the following webpage
![www.genetics.ucla.edu/labs/horvath/CoexpressionNetwork]

Download the zip file containing:

1) R function file: "NetworkFunctions.txt", which contains several R functions
    needed for Network Analysis.
    
2) The data file "gbm55old_dchip_14kALL_cox_8000mvgenes2.csv "

3) Of course, this file: "GBMTutorialHorvath.txt"

(I already did this)

```{r}
source("NetworkFunctions.R")
# read in the R libraries
library(MASS) # standard, no need to install
library(class) # standard, no need to install
library(cluster)
#library(impute) # install it for imputing missing value 
library(WGCNA)
options(stringsAsFactors = F)
library(gplots)
```

## read in the 8000 most varying genes (GBM microarray data)

```{r}
dat0=read.csv("gbm55old_dchip_14kALL_cox_8000mvgenes2.csv")
# this contains information on the genes
datSummary=dat0[,1:9] 
```

The following data frame contains the gene expression data: columns are genes, rows are arrays (samples)

```{r}
datExpr = t(dat0[,10:64])

no.samples = dim(datExpr)[[1]]
dim(datExpr)
rm(dat0);gc()
```


# To choose a cut-off value, 

we propose to use the Scale-free Topology Criterion (Zhang and Horvath 2005). Here the focus is on the linear regression model fitting index (denoted below by scale.law.R.2) that quantify the extent of how well a network satisfies a scale-free topology. The function PickSoftThreshold can help one to estimate the cut-off value when using hard thresholding with the step adjacency function. The first column (different from the row numbers) lists the soft threshold Power The second column reports the resulting scale free topology fitting index R^2 (scale.law.R.2) The third column reports the slope of the fitting line. The fourth column reports the fitting index for the truncated exponential scale free model# Usually we ignore it. The remaining columns list the mean, median and maximum connectivity. To a soft threshold (power) with the scale free topology criterion: aim for reasonably high scale free R^2 (column 2), higher than say .80 and negative slope (around -1, col 4). In practice, we pick the threshold by looking for a "kink" in the relationship between R^2 and power, see below. 

## Soft thresholding

Now we investigate soft thesholding with the power adjacency function

```{r}
powers1=c(seq(1,10,by=1),seq(12,20,by=2))
RpowerTable=pickSoftThreshold(datExpr, powerVector=powers1)[[2]] 

gc()
cex1=0.7
par(mfrow=c(1,2))
plot(RpowerTable[,1], -sign(RpowerTable[,3])*RpowerTable[,2],xlab="
Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n")
text(RpowerTable[,1], -sign(RpowerTable[,3])*RpowerTable[,2],
labels=powers1,cex=cex1,col="red")
# this line corresponds to using an R^2 cut-off of h
abline(h=0.95,col="red")
plot(RpowerTable[,1], RpowerTable[,5],xlab="Soft Threshold (power)",ylab="Mean
Connectivity", type="n")
text(RpowerTable[,1], RpowerTable[,5], labels=powers1, cex=cex1,col="red") 
```

Note that at power=6, the curve has an elbow or kink, i.e. for this power the scale free topology fit does not improve after increasing the power. This is why we choose beta1=6
Also the scale free topology criterion with a R^2 threshold of 0.95 would lead us to pick a power of 6.
Note that there is a natural trade-off between maximizing scale-free topology model fit (R^2) and
maintaining a high mean number of connections: parameter values that lead to an R^2 value close
to 1 may lead to networks with very few connections. Actually, we consider a signed
version of the scale free topology fitting index. Since it is biologically implausible that a networks
contains more hub genes than non-hub genes, we multiply R^2 with -1 if the slope of
the regression line between log_{10}(p(k)) and log_{10}(k) is positive. 

These considerations motivate us to propose the following {scale-free topology criterion} for
choosing the parameters of an adjacency function: Only consider those parameter values
that lead to a network satisfying scale-free topology at least approximately, e.g. signed R^2>0.80.
In addition, we recommend that the user take the following additional considerations into
account when choosing the adjacency function parameter. First, the mean connectivity should be
high so that the network contains enough information (e.g. for module detection). Second, the slope
of the regression line should be around -1.
When considering the power adjacency functions, we find the relationship between R^2 and the
adjacency function parameter (beta) is characterized by a saturation curve type of. In our
applications, we use the first parameter value where saturation is reached as long
as it is above 0.8.
Below we study how the biological findings depend on the choice of the power. 

```{r}
beta1=6
Connectivity=softConnectivity(datExpr,power=beta1)-1
```


## Let’s create a scale free topology plot.
The black curve corresponds to scale free topology and the red curve corresponds to truncated scale free topology.

```{r}
par(mfrow=c(1,1))
scaleFreePlot(Connectivity, main=paste("soft threshold, power=",beta1), truncated=F); 
```

# Module Detection
An important step in network analysis is module detetion. Here we use methods that use clustering in combination with the topological
 overlap matrix.
 This code allows one to restrict the analysis to the most connected genes,
 which may speed up calculations when it comes to module detection.

```{r}
ConnectivityCut = 3600 # number of most connected genes that will be considered

# Incidentally, in the paper by Mischel et al (2005) we considered all 3600 #genes.
ConnectivityRank = rank(-Connectivity)
restConnectivity = ConnectivityRank <= ConnectivityCut

# thus our module detection uses the following number of genes
sum(restConnectivity)

# Now we define the adjacency matrix for the 3600 most connected genes
ADJ= adjacency(datExpr[,restConnectivity],power=beta1)
gc()

# The following code computes the topological overlap matrix based on the
# adjacency matrix.
# TIME: This about a few minutes....
dissTOM=TOMdist(ADJ)
gc() 
```


Now we carry out hierarchical clustering with the TOM matrix.

This takes a couple of minutes.

```{r}
hierTOM = hclust(as.dist(dissTOM),method="average");
par(mfrow=c(1,1))
plot(hierTOM,labels=F) 
```

According to our definition, modules correspond to branches of the tree. The question is what height cut-off should be used? This depends on the biology. Large heigth values lead to big modules, small values lead to small but tight modules. In reality, the user should use different thresholds to see how robust the findings are. 

The function cutreeStatistColor colors each gene by the branches that result from choosing a particular height cut-off. GREY IS RESERVED to color genes that are not part of any module. We only consider modules that contain at least 125 genes.

```{r}
colorh1= cutreeStaticColor(hierTOM,cutHeight = 0.94, minSize = 125)
# The above should be identical to colorh1=datSummary$color1[restConnectivity]
par(mfrow=c(2,1),mar=c(2,4,1,1))
plot(hierTOM, main="Cluster Dendrogram", labels=F, xlab="", sub="");
plotColorUnderTree(hierTOM,colors=data.frame(module=colorh1))
title("Module (branch) color") 
```

### COMMENTS:

1) The colors are assigned based on module size. Turquoise (others refer to it as cyan) colors
the largest module, next comes blue, next brown, etc. Just type table(colorh1) to figure out
which color corresponds to what module size.

2) The minimum module size (minsize1=125) is unusually large. As default, we recommend
minsize1=50.

3) Here we choose a fixed height cut-off (h1) for cutting off branches. But we have also
developed a more flexible method for cutting off branches that adaptively choose a different
height for each branch. The resulting dynamic tree cutting algorithm (cutreeDynamic) is
desccribed in Langfelder et al (2008). 

An alternative view of this is the so called TOM plot that is generated by the function TOMplot

Inputs: TOM distance measure, hierarchical (hclust) object, color

Warning: for large gene sets, say more than 2000 genes this will take a while. I recommend you skip this.

```{r}
TOMplot(dissTOM , hierTOM, colorh1) 
```

We also propose to use classical multi-dimensional scaling plots for visualizing the network. Here we chose 2 scaling dimensions. This also takes about 10 minutes...

```{r}
cmd1=cmdscale(as.dist(dissTOM),2)
par(mfrow=c(1,1))
plot(cmd1, col=as.character(colorh1), main="MDS plot",xlab="Scaling Dimension
1",ylab="Scaling Dimension 2") 
```

# Module significance

Next we define a gene significance variable as minus log10 of the univarite Cox regression pvalue for predicting survival on the basis of the gene epxression info

```{r}
# this defines the gene significance for all genes
GeneSignificanceALL=-log10(datSummary$pCox)
# gene significance restricted to the most connected genes:
GeneSignificance=GeneSignificanceALL[restConnectivity]
```


The function verboseBarplot creates a bar plot that shows whether modules are enriched with essential genes. It also reports a Kruskal Wallis P-value. The gene significance can be a binary variable or a quantitative variable. It also plots the 95% confidence interval of the mean

```{r}
par(mfrow=c(1,1))
verboseBarplot(GeneSignificance,colorh1,main="Module Significance ",
col=levels(factor(colorh1)) ,xlab="Module" ) 
```

Note that the brown module have a high mean value of gene significance.
As aside for the experts, we should mention that the p-value (Kruskal Wallis test) cannot be trusted
due to dependence between the genes. The p-value should really be interpreted as a descriptive (not
inferential) measure. 

To get a sense of how related the modules are one can summarize each module by its first eigengene (referred to as principal components) and then correlate these module eigengenes with each other. 

```{r}
datME=moduleEigengenes(datExpr[,restConnectivity],colorh1)[[1]]
```

We define a dissimilarity measure between the module eigengenes that keeps track of the sign of the correlation between the module eigengenes.

```{r}
dissimME=1-(t(cor(datME, method="p")))/2
hclustdatME=hclust(as.dist(dissimME), method="average" )
par(mfrow=c(1,1))
plot(hclustdatME, main="Clustering tree based on the module eigengenes of modules")
```

Now we create scatter plots of the samples (arrays) along the module eigengenes.

```{r}
datMEordered=datME[,hclustdatME$order]
pairs( datMEordered, upper.panel = panel.smooth, lower.panel = panel.cor,
diag.panel=panel.hist ,main="Relation between module eigengenes") 
```

Message: the module eigengenes (first PC) of different modules may be highly correlated.
WGCNA can be interpreted as a biologically motivated data reduction scheme that allows for
dependency between the resulting components. Compare this to principal component analysis that
would impose orthogonality between the components.
Since modules may represent biological pathways there is no biological reason why modules
should be orthogonal to each other.
Aside: If you are interested in networks comprised of module eigengenes, the following article may
be of interest:
Langfelder P, Horvath S (2007) Eigengene networks for studying the relationships between coexpression modules. BMC Systems Biology 2007, 1:54 

To study how connectivity is related to mean gene expression or variance of gene expression we create the following plot.

```{r}
mean1=function(x) mean(x,na.rm=T)
var1=function(x) var(x,na.rm=T)
meanExpr=apply( datExpr[,restConnectivity],2,mean1)
varExpr=apply( datExpr[,restConnectivity],2,var1)
par(mfrow=c(1,2))
plot(Connectivity[restConnectivity],meanExpr, col=as.character(colorh1),
main="Mean(Expression) vs K",xlab="Connectivity")
plot (Connectivity[restConnectivity],varExpr, col= as.character(colorh1), main="Var(Expression)
vs K" ,xlab="Connectivity") 
```

The following produces heatmap plots for each module.

Here the rows are genes and the columns are samples.

Well defined modules results in characteristic band structures since the corresponding genes are highly correlated.

```{r}
par(mfrow=c(2,1), mar=c(1, 2, 4, 1))
ClusterSamples=hclust(dist(datExpr[,restConnectivity] ),method="average")
# for the first (turquoise) module we use
which.module="turquoise"
rg=colorpanel(50, low="green", mid="black", high="red")
heatmap(t(scale(datExpr[ClusterSamples$order,restConnectivity][,colorh1==which.module ])),col=rg,Colv=NA)
# for the second (blue) module we use
which.module="blue"
heatmap(t(scale(datExpr[ClusterSamples$order,restConnectivity][,colorh1==which.module ])),col=rg,Colv=NA)
```

Now we extend the color definition to all genes by coloring all non-module genes grey.

```{r}
color1=rep("grey",dim(datExpr)[[2]])
color1[restConnectivity]=as.character(colorh1)
# The function intramodularConnectivity computes the whole network connectivity kTotal,
# the within module connectivity (kWithin). kOut=kTotal-kWithin and
# and kDiff=kIn-kOut=2*kIN-kTotal
ConnectivityMeasures=intramodularConnectivity(ADJ,colors=colorh1)
names(ConnectivityMeasures)

# The following plots show the gene significance vs intramodular connectivity
colorlevels=levels(factor(colorh1))
par(mfrow=c(2,3),mar=c(5, 4, 4, 2) + 0.1)
for (i in c(1:length(colorlevels) ) ) {
whichmodule=colorlevels[[i]];restrict1=colorh1==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh1[restrict1],main= paste("set I,",
whichmodule),ylab="Gene Significance",xlab="Intramodular k")
} 
```

Generalizing the intramodular connectivity measure to *all* genes.
Note that the intramodular connectivity measure is only defined for the genes inside a given
module. But in practice it can be very important to measure how connected a given genes is to
biologically interesting modules.
Toward this end, we define a module eigengene based connectivity measure for each gene as the
correlation between a the gene expression and the module eigengene.
Specifically,

kMEbrown(i)=cor(x(i), PCbrown)

where x(i) is the gene expression profile of the i-th gene and PCbrown is the module eigengene of
the brown module. Note that the definition does not require that the i-th gene is a member of the
brown module. 

The following data frame contains the kME values corresponding to each module.

```{r}
datKME=signedKME(datExpr, datME)
#Note we have an intramodular connectivity measure for each color.
names(datKME)
```

Note that the intramodular connectivity has been computed for each of the 8000 genes.

```{r}
dim(datKME)
attach(datKME)
```

Question: how do the kME measure relate to the standard intramodular connectivity?

```{r}
whichmodule="brown"
restrictGenes= colorh1== whichmodule
par(mfrow=c(1,1))
verboseScatterplot(ConnectivityMeasures$kWithin[ restrictGenes],
(datKME$kMEbrown[restConnectivity][restrictGenes])^beta1 ,xlab="kIN",ylab="kME^power",
col=whichmodule,main="Relation between two measures of intramodular k, ") 
```

Note that after raising kME to a power of 6, it is highly correlated with kWithin. A theoretical
derivation of this finding can be found in Horvath and Dong (2008). 

Question: find genes with high gene significance (Cox-pvalue smaller than 0.05) and high
intramodular connectivity in the brown module.

```{r}
attach(datKME)
FilterGenes= GeneSignificanceALL > -log10(0.05) & abs(kMEbrown)>.85
table(FilterGenes)
datSummary[FilterGenes,] 
```

Comments:
The ASPM gene colored in red was the focus of the paper Horvath et al (2006) but there are many
other interesting genes. 

To illustrate the use of the kME measures, we also address the following questions


Question: Screen for significant genes that have a negative correlation with the brown module
eigengene

```{r}
FilterGenes= GeneSignificanceALL> -log10(0.05) & -kMEbrown> .5 # notice the red minus sign!
table(FilterGenes)
datSummary[FilterGenes,] 
```

Question: Screen for significant genes that are close to the brown module and the green module
and far away from the yellow module. Answer

```{r}
FilterGenes= GeneSignificanceALL > -log10(0.05) & abs(kMEbrown)>.5 & abs(kMEgreen)>.5
table(FilterGenes)
datSummary[FilterGenes,]
```

Question: Screen for significant genes that are close to the brown module and far away from the
yellow module. Answer

```{r}
FilterGenes= GeneSignificanceALL > -log10(0.05) & abs(kMEbrown)>.6 & abs(kMEyellow)<.3
table(FilterGenes) 
```


# How to output the data? 

```{r}
datout=data.frame(datSummary, colorNEW=color1, ConnectivityNew=Connectivity,datKME )
write.table(datout, file="OutputCancerNetwork.csv", sep=",", row.names=F)
```

# Robustness with regard to the soft threshold

We find that the results of weighted gene co-expression network analysis are highly robust with
regard to the soft threshold beta. Here we show some results that demonstrate this point.
Now we want to see how the correlation between kWithin and gene significance changes for
different SOFT thresholds (powers). This analysis is restricted to the brown module genes.

Also we compare the 2 different connectivity measures: The standard connectivity measure is
defined as the row sum of the adjacency matrix. The non-standard connectivity measure
(kTOM.IN) is defined as row sum of the topological overlap matrix .

Now we want to see how the correlation between kWithin and gene significance changes for different powers beta within the BROWN module.

```{r}
corhelp=cor(datExpr[,restConnectivity],use="pairwise.complete.obs")
whichmodule="brown"
datconnectivitiesSoft=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(powers1)))
names(datconnectivitiesSoft)=paste("kWithinPower",powers1,sep="")
for (i in c(1:length(powers1)) ) {
datconnectivitiesSoft[,i]=apply(abs(corhelp[colorh1==whichmodule,
colorh1==whichmodule])^powers1[i],1,sum)}
SpearmanCorrelationsSoft=signif(cor(GeneSignificance[ colorh1==whichmodule],
datconnectivitiesSoft, method="s",use="p"))
# Here we use the new connectivity measure based on the topological overlap matrix
datKTOM.IN=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(powers1)))
names(datKTOM.IN)=paste("omegaWithinPower",powers1,sep="")
for (i in c(1:length(powers1)) ) {
datconnectivitiesSoft[,i]=apply(
1-TOMdist(abs(corhelp[colorh1==whichmodule, colorh1==whichmodule])^powers1[i])
,1,sum)}
SpearmanCorrelationskTOMSoft=as.vector(signif(cor(GeneSignificance[ colorh1==whichmodule],
datconnectivitiesSoft, method="s",use="p")))
par(mfrow=c(1,1), mar=c(5, 4, 4, 2) +0.1)
plot(powers1, SpearmanCorrelationsSoft, main="Cor(Connectivity,Gene Significance) vs Soft
Thresholds(powers)",ylab="Spearman Correlation(Gene Significance,k.in)",xlab="Power
beta",type="n",ylim=range(c(SpearmanCorrelationsSoft,
SpearmanCorrelationskTOMSoft),na.rm=T)
)
text(powers1, SpearmanCorrelationsSoft,labels=powers1,col="red")
# this draws a vertical line at the tau that was chosen by the
# scale free topology criterion.
abline(v=6,col="red") 

points(powers1, SpearmanCorrelationskTOMSoft, type="n")
text(powers1, SpearmanCorrelationskTOMSoft,labels=powers1,col="orange") 
```

Now we define the intramodular clustering coefficient (also see the section below)

```{r}
datCCinSoft=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(powers1)))
names(datCCinSoft)=paste("CCinSoft",powers1,sep="")
for (i in c(1:length(powers1)) ) {
datCCinSoft[,i]= clusterCoef(abs(corhelp[colorh1==whichmodule,
colorh1==whichmodule])^powers1[i])
}
SpearmanCorrelationsCCinSoft=as.vector(signif(cor(GeneSignificance[ colorh1==whichmodule],
datCCinSoft, method="s",use="p")))
dathelpSoft=data.frame(signedRsquared=-sign(RpowerTable[,3])*RpowerTable[,2], corGSkINSoft
=as.vector(SpearmanCorrelationsSoft), corGSwINSoft=
as.vector(SpearmanCorrelationskTOMSoft),corGSCCSoft=as.vector(SpearmanCorrelationsCCinSoft))
matplot(powers1,dathelpSoft,type="l",lty=1,lwd=3,col=c("black","red","blue","green"),ylab="",xlab="beta")
abline(v=6,col="red")
legend(13,0.5, c("signed R^2","r(GS,k.IN)","r(GS,kTOM.IN)","r(GS,cc.in)"),
col=c("black","red","blue","green"), lty=1,lwd=3,ncol = 1, cex=1)
```

## Comment: 
the intramodular cluster coefficient (green line) achieves the highest correlation
with the gene significance. The TOM based intramodular connectivity kTOM.in (blue line) is
superior to the standard connectivity measure k.in (red line) in this application.
The vertical line corresponds to the power picked by the scale free topology criterion.
The scale free topology criterion leads to near optimal biological signal when using kTOM.IN.


CAVEAT: It is worth mentioning that in other real data sets k.in outperforms cc.in and kTOM.IN.

## Computation of the cluster coefficient in the weighted network.

The clustering coefficient measures the cliquishness of a gene. Many references use this concept. For our definition of the clustering coefficient in weighted networks consult Zhang and Horvath
(2005) and Dong and Horvath (2007). 

Here we study how the clustering coefficient depends on the connectivity.

Since this is computationally intensive (around 15 minutes), we recommend to skip it.
```{r}
CC= clusterCoef(ADJ)
gc()
```

Now we plot cluster coefficient versus connectivity for all genes

```{r}
par(mfrow=c(1,1),mar=c(2,2,2,1))
plot(Connectivity[restConnectivity],CC,col=as.character(colorh1),xlab="Connectivity",ylab="Clust
er Coefficient") 
```

This compute the correlation between cluster coefficient and connectivity within each module.

```{r}
restHub= Connectivity[restConnectivity]>0
by(data.frame(CC=CC[restHub], k=Connectivity[restConnectivity][restHub]),
INDICES=colorh1[restHub],FUN=cor) 
```


# Unweighted analysis


To construct an unweighted network (hard thresholding), we consider the following vector of potential thresholds. 

```{r}
thresholds1= c(seq(.1,.5, by=.1), seq(.55,.95, by=.05) ) 
```

To choose a cut-off value, we propose to use the Scale-free Topology Criterion (Zhang and Horvath 2005). Here the focus is on the linear regression model fitting index (denoted below by scale.law.R.2) that quantify the extent of how well a network satisfies a scale-free topology. The function PickHardThreshold can help one to estimate the cut-off value when using hard thresholding with the step adjacency function. The first column lists the threshold ("cut"), the second column lists the corresponding p-value based on the Fisher transform. The third column reports the resulting scale free topology fitting index R^2. The fourth column reports the slope of the fitting line. The fifth column reports the fitting index for the truncated exponential scale free model. Usually we ignore it. The remaining columns list the mean, median and maximum connectivity. To pick a hard threshold (cut) with the scale free topology criterion: aim for high scale free R^2 (column 3), high connectivity (col 6) and negative slope (around -1, col 4).

```{r}
RdichotTable=WGCNA::pickHardThreshold(datExpr, cutVector=thresholds1)[[2]]
gc()
```


Let’s plot the scale free topology model fitting index (R^2) versus the cut-off tau. However, the R^2 values of those cut-offs that lead to a negative slope have been pre-multiplied by -1. 

```{r}
cex1=0.7
gc()
par(mfrow=c(1,2))
plot(RdichotTable[,1], -sign(RdichotTable[,4])*RdichotTable[,3],xlab="Hard Threshold
tau",ylab="Scale Free Topology Model Fit,signed R^2", type="n")
text(RdichotTable[,1], -sign(RdichotTable[,4])*RdichotTable[,3] , labels=thresholds1,cex=cex1)
# this line corresponds to using an R^2 cut-off of h
abline(h=0.95,col="red")
plot(RdichotTable[,1], RdichotTable[,6],xlab="Hard Threshold tau",ylab="Mean Connectivity",
type="n")
text(RdichotTable[,1], RdichotTable[,6] , labels=thresholds1, cex=cex1) 
```

To choose a cut-off value tau, we propose to use the Scale-free Topology Criterion (Zhang and
Horvath 2005). Here the focus is on the linear regression model
fitting index (denoted as scale.law.R.2) that quantify the extent of how well
a network satisfies a scale-free topology. We choose the cut value (tau) of 0.7 for the correlation
matrix since this is where the R^2 curve seems to saturates. The red line corresponds to R^2= 0.95.
From the above table, we find that the resulting slope looks OK (negative and around -1), and the
mean number of connections looks good Below we investigate different choices of tau.

Now we want to see how the correlation between kWithin and gene significance changes for different hard threshold values tau within the BROWN module. 

```{r}
corhelp=cor(datExpr[,restConnectivity],use="pairwise.complete.obs")
whichmodule="brown"
# the following data frame contains the intramodular connectivities
# corresponding to different hard thresholds
datconnectivitiesHard=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(thresholds1)))
names(datconnectivitiesHard)=paste("kWithinTau",thresholds1,sep="")
for (i in c(1:length(thresholds1)) ) {
datconnectivitiesHard[,i]=apply(abs(corhelp[colorh1==whichmodule,
colorh1==whichmodule])>=thresholds1[i],1,sum)}
SpearmanCorrelationsHard=signif(cor(GeneSignificance[ colorh1==whichmodule],
datconnectivitiesHard, method="s",use="p")) 
```


Now we define the new connectivity measure omega based on the TOM matrix. It simply considers TOM as adjacency matrix... 

```{r}
datkTOMINHard=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(thresholds1)))
names(datkTOMINHard)=paste("omegaWithinHard",thresholds1,sep="")
for (i in c(1:length(thresholds1)) ) {
datconnectivitiesHard[,i]=apply(
1-TOMdist(apply(abs(corhelp[colorh1==whichmodule,
colorh1==whichmodule])>thresholds1[i],2,as.numeric)),1,sum)}
SpearmanCorrelationskTOMHard=as.vector(signif(cor(GeneSignificance[
colorh1==whichmodule], datconnectivitiesHard, method="s",use="p")))
```

Now we compare the performance of the 2 connectivity measures (k.in and kTOM.IN) across different hard thresholds when it comes to predicting prognostic genes in the brown module 

```{r}
par(mfrow=c(1,1), mar=c(5, 4, 4, 2) +0.1)
plot(thresholds1, SpearmanCorrelationsHard, main="
Cor(Connectivity,Gene Significance) vs Hard Thresholds",ylab="Spearman Correlation(Gene
Significance,Connectivity)",xlab="Threshold tau", type="n",
ylim=range(c(SpearmanCorrelationsHard, SpearmanCorrelationskTOMHard),na.rm=T))
text(thresholds1, SpearmanCorrelationsHard,labels=thresholds1,col="black")
# this draws a vertical line at the tau that was chosen by the
# scale free topology criterion.
abline(v=0.7,col="red")
points(thresholds1, SpearmanCorrelationskTOMHard, type="n")
text(thresholds1, SpearmanCorrelationskTOMHard,labels=thresholds1,col="blue") 
```

Now we define the intramodular cluster coefficient 

```{r}
datCCinHard=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(thresholds1
)))
names(datCCinHard)=paste("CCinHard",thresholds1,sep="")
for (i in c(1:length(thresholds1)) ) {
datCCinHard[,i]= clusterCoef(apply(abs(corhelp[colorh1==whichmodule,
colorh1==whichmodule])>thresholds1[i],2,as.numeric))}
SpearmanCorrelationsCCinHard=as.vector(signif(cor(GeneSignificance[ colorh1==whichmodule],
datCCinHard, method="s",use="p"))) 
```

Now we compare the performance of the connectivity measures (k.in, kTOM.IN, cluster coefficience) across different hard thresholds when it comes to predicting prognostic genes in the brown module

```{r}
dathelpHard=data.frame(signedRsquared=-sign(RdichotTable[,4])*RdichotTable[,3],
corGSkINHard =as.vector(SpearmanCorrelationsHard), corGSwINHard=
as.vector(SpearmanCorrelationskTOMHard),corGSCCHard=as.vector(SpearmanCorrelationsCCinHard))
matplot(thresholds1,dathelpHard,type="l",lty=1,lwd=3,col=c("black","red","blue","green"),ylab=""
,xlab="tau",xlim=c(.2,1))
legend(0.75,0, c("signed R^2","r(GS,k.in)","r(GS,kTOM.IN)","r(GS,cc.in)"),
col=c("black","red","blue","green"), lty=1,lwd=3,ncol = 1, cex=1)
abline(v=.7,col="red") 
```

Note that very high or very small threshold values lead to a small correlation, i.e. a diminished biological signal. The red line corresponds to the threshold that was picked using the scale free topology criterion. The scale free topology criterion picked a threshold that leads to very high significant correlation between node connectivity and gene significance.

```{r}
AdjMatHARD=abs(cor(datExpr[,restConnectivity]))>0.70+0.0
diag(AdjMatHARD)=0
cluster.coefrestHARD= clusterCoef(AdjMatHARD)
ConnectivityHARD= apply(AdjMatHARD,2,sum)
par(mfrow=c(1,1))
plot(ConnectivityHARD,cluster.coefrestHARD,col=as.character(colorh1),xlab="Connectivity",ylab="Cluster Coefficient" ) 
```

Now we correlate the cluster coefficient with connectivity by module in the unweighted network 

```{r}
restHub=ConnectivityHARD>100
by(data.frame(CC= cluster.coefrestHARD[restHub], k=ConnectivityHARD[restHub]),
INDICES=colorh1[restHub],FUN=cor) 
```

# Let’s compare a summary of soft thresholding to one of hard thresholding 

```{r}
apply(dathelpHard,2,summary)
```

• Using the Max. values, we find that when it comes to correlating gene significance with a
centrality measure (connectivity or cluster coefficient), the soft intramodular cluster
coefficient is most highly correlated with gene significance. Next comes the soft TOM
based connectivity (wIN), then kIN. In other applications, the cluster coefficient is not the
best centrality measure. But please let us know if you find empirical evidence that the
cluster coefficient is a good centrality measure. In the latest version of the manuscript
Zhang and Horvath (2005) we present a theoretical argument that shows that is a weak
positive correlation between intramodular cluster coefficient and intramodular connectivity
in weighted networks. In contrast, one finds a negative correlation between cluster
coefficient and connectivity in unweighted networks (see the plot in the appendix).
• We find that soft thresholding is superior to hard thresholding especially for low values of
the scale free topology R^2.
• In our opinion, soft centrality (connectivity) measures are better than hard measures
because they are relatively robust with respect to the parameter of the adjacency function.
For soft thresholding even choosing a power of beta=1 leads to a good biological signal
(correlation). In contrast, choosing a hard threshold of tau=0.2 leads to a much reduced
biological signal. Robustness is a very attractive property in this type of analysis since
picking parameters of the adjacency function is rather ad-hoc.
• The Scale free topology criterion leads to estimates of the adjacency function that often
have good biological signal. 


# APPENDIX: Constructing an unweighted networks and comparing it to the weighted nework.

Here we study whether the `soft’ modules of the unweighted network described above can also be
found in the unweighted network

Recall that the soft module assignment in the 3600 most connected genes is given by 

```{r}
colorh1=as.character(datSummary$color1[restConnectivity])
```

Let’s define the adjacency matrix of an unweighted network 

```{r}
ADJ= abs(cor(datExpr[,restConnectivity],use="p"))>0.7
ADJ <- apply(ADJ,2,as.numeric)
gc()
# This is the unweighted connectivity
k=as.vector(apply(ADJ,2,sum)) 
```

Let’s compare weighted to unweighted connectivity in a scatter plot

```{r}
plot(k, Connectivity[restConnectivity],xlab="Unweighted
Connectivity",ylab="Weighted Connectivity",main=paste( "correlation =
",signif(cor(k,Connectivity[restConnectivity]),2)),col=colorh1)
```

Comments:
• the connectivity measures is highly preserved between weighted and unweighted networks
as long as the scale free topology criterion is used for network construction. It is re-assuring
that the biological findings will be robust with respect to the network construction method
• The genes with the highest whole network connectivity are contained in the turquoise
module, which happens to be the largest module. The second most connected genes are in
the blue module, which is the second largest module, etc.

The following code computes the topological overlap matrix based on the adjacency matrix

TIME: Takes about 10 minutes.... 

```{r}
dissTOM=TOMdist(ADJ)
gc() 
```

Now we carry out hierarchical clustering with the TOM matrix. Branches of the resulting clustering tree will be used to define gene modules. 

```{r}
hierTOM = hclust(as.dist(dissTOM),method="average");
par(mfrow=c(1,1))
plot(hierTOM,labels=F) 
```

By our definition, modules correspond to branches of the tree. The function modulecolor2 colors each gene by the branches that result from choosing a particular height cut-off. GREY IS RESERVED to color genes that are not part of any module. We only consider modules that contain at least 125 genes.

```{r}
colorh2=as.character(modulecolor2(hierTOM,h1=.75, minsize1=125)) 
```

The following table shows that there is fairly high agreement between the soft and the hard module assignments 

```{r}
table(colorh1,colorh2)

#Rand index to measure agreement between the clusterings
randIndex(table(colorh1,colorh2)) 
```

Note that the brown module in the weighted network (colorh1) is a little bit larger than the corresponding module in the unweighted network. But the point is that it is highly preserved. Since this module is of biological interest, the good news is that the biological findings are robust with respect to the network construction method as long as the scale free topology criterion is used to construct the network. 

```{r}
par(mfrow=c(3,1), mar=c(2,2,2,1))
plot(hierTOM, main="Unweighted Network Module Tree ", labels=F, xlab="",
sub="");
hclustplot1(hierTOM, colorh2, main="Colored by Unweighted modules")
hclustplot1(hierTOM, colorh1, main="Colored by Weighted modules")
```

An alternative view of this is the so called TOM plot that is generated by the function TOMplot 

Inputs: TOM distance measure, hierarchical (hclust) object, color

Here we use the unweighted module tree but color it by the weighted modules.

```{r}
TOMplot(dissTOM , hierTOM, as.character(datSummary$color1[restConnectivity]))
gc() 
```

Comment: module assignment is highly preserved. 

```{r}
verboseBarplot(GeneSignificance,colorh1,main="Gene Significance in Soft Modules") 
verboseBarplot(GeneSignificance,colorh2,main="Gene Significance in Hard Modules") 
```


Now we compute the cluster coefficient. Since this is computationally intensive, we recommend to skip it. 

```{r}
CC= clusterCoef(ADJ)
gc() 

plot(k,CC,col= as.character(datSummary$color1[restConnectivity]),xlab="Connectivity
(Hard)",ylab="Cluster Coefficient") 
```

Comment: for unweighted networks there is an inverse relationship between cluster coefficient and connectivity. This is different from the case of weighted networks. In our opinion, this inverse relationship is an artifact of hard thresholding, see Zhang and Horvath (2005). 

Now we study how intramodule connectivity relates to gene significance 

```{r}
ConnectivityMeasures=intramodularConnectivity(ADJ,colorh2) 
```

The following plots would show the gene significance vs intromodular connectivity

```{r}
par(mfrow=c(2,3))
colorlevels=unique(colorh2)
whichmodule=colorlevels[[1]];restrict1=colorh2==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh2[restrict1],main=whichmodule)
whichmodule= colorlevels[[2]];restrict1=colorh2==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh2[restrict1],main=whichmodule)
whichmodule= colorlevels[[3]];restrict1=colorh2==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh2[restrict1],main=whichmodule)
whichmodule= colorlevels[[4]];restrict1=colorh2==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh2[restrict1],main=whichmodule)
whichmodule= colorlevels[[5]];restrict1=colorh2==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh2[restrict1],main=whichmodule)
whichmodule= colorlevels[[6]];restrict1=colorh2==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh2[restrict1],main=whichmodule)
```