---
title: "Original Analysis"
author: "Sam Ozminkowski"
date: "2023-05-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
setwd("/Users/samoz/Documents/phd/year1/stat 465/project/STAT_465_gene_coexpression/analysis/")
```

```{r}
source("NetworkFunctions.R")

library(MASS) # standard, no need to install
library(class) # standard, no need to install
library(cluster)
library(WGCNA)
options(stringsAsFactors = F)
library(gplots)
```

This is basically just the tutorial but i trimmed a bunch of the fat

## read in the 8000 most varying genes (GBM microarray data)

```{r}
dat0=read.csv("data/gbm55old_dchip_14kALL_cox_8000mvgenes2.csv")
# this contains information on the genes
datSummary=dat0[,1:9]
datExpr = t(dat0[,10:64])

no.samples = dim(datExpr)[[1]]
dim(datExpr)
rm(dat0);gc()
```

# Soft Clustering

```{r}
library(ggplot2)

powers1=c(seq(1,10,by=1),seq(12,20,by=2))
RpowerTable=pickSoftThreshold(datExpr, powerVector=powers1)[[2]] 

gc()
cex1=0.7
par(mfrow=c(1,2))
plot(RpowerTable[,1], -sign(RpowerTable[,3])*RpowerTable[,2],
     xlab="Soft Threshold (power)",
     ylab="Scale Free Topology Model Fit,signed R^2",type="n")

text(RpowerTable[,1], -sign(RpowerTable[,3])*RpowerTable[,2],
     labels=powers1,cex=cex1,col="red")
# this line corresponds to using an R^2 cut-off of h
abline(h=0.95,col="red")

plot(RpowerTable[,1], RpowerTable[,5],xlab="Soft Threshold (power)",ylab="Mean
Connectivity", type="n")
text(RpowerTable[,1], RpowerTable[,5], labels=powers1, cex=cex1,col="red") 
```


```{r}
beta1=6
Connectivity=softConnectivity(datExpr,power=beta1)-1
```

## Let’s create a scale free topology plot.
The black curve corresponds to scale free topology and the red curve corresponds to truncated scale free topology.

```{r}
par(mfrow=c(1,1))
scaleFreePlot(Connectivity, main=paste("soft threshold, power=",beta1), truncated=F)
```


## Module Detection

```{r}
ConnectivityCut = 3600 # number of most connected genes that will be considered

# Incidentally, in the paper by Mischel et al (2005) we considered all 3600 #genes.
ConnectivityRank = rank(-Connectivity)
restConnectivity = ConnectivityRank <= ConnectivityCut

# thus our module detection uses the following number of genes
sum(restConnectivity)

# Now we define the adjacency matrix for the 3600 most connected genes
ADJ= adjacency(datExpr[,restConnectivity],power=beta1)
gc()

# The following code computes the topological overlap matrix based on the
# adjacency matrix.
# TIME: This about a few minutes....
dissTOM=TOMdist(ADJ)
gc() 

# heirarchical clustering
hierTOM = hclust(as.dist(dissTOM),method="average");
par(mfrow=c(1,1))
plot(hierTOM,labels=F) 
```

```{r}
colorh1= cutreeStaticColor(hierTOM,cutHeight = 0.94, minSize = 125)
# The above should be identical to colorh1=datSummary$color1[restConnectivity]
par(mfrow=c(2,1),mar=c(2,4,1,1))
plot(hierTOM, main="Cluster Dendrogram", labels=F, xlab="", sub="");
plotColorUnderTree(hierTOM,colors=data.frame(module=colorh1))
title("Module (branch) color") 
```

```{r}
TOMplot(dissTOM , hierTOM, colorh1) 
```

## Module significance

```{r}
# this defines the gene significance for all genes
GeneSignificanceALL=-log10(datSummary$pCox)
# gene significance restricted to the most connected genes:
GeneSignificance=GeneSignificanceALL[restConnectivity]
```

The p-value (Kruskal Wallis test) cannot be trusted due to dependence between the genes. 

To get a sense of how related the modules are one can summarize each module by its first eigengene (referred to as principal components) and then correlate these module eigengenes with each other. 

```{r}
datME=moduleEigengenes(datExpr[,restConnectivity],colorh1)[[1]]
```

We define a dissimilarity measure between the module eigengenes that keeps track of the sign of the correlation between the module eigengenes.

```{r}
dissimME=1-(t(cor(datME, method="p")))/2
hclustdatME=hclust(as.dist(dissimME), method="average" )
par(mfrow=c(1,1))
plot(hclustdatME, main="Clustering tree based on the module eigengenes of modules")
datMEordered=datME[,hclustdatME$order]
pairs( datMEordered, upper.panel = panel.smooth, lower.panel = panel.cor,
diag.panel=panel.hist ,main="Relation between module eigengenes") 
```


To study how connectivity is related to mean gene expression or variance of gene expression we create the following plot.

```{r}
mean1=function(x) mean(x,na.rm=T)
var1=function(x) var(x,na.rm=T)
meanExpr=apply( datExpr[,restConnectivity],2,mean1)
varExpr=apply( datExpr[,restConnectivity],2,var1)
par(mfrow=c(1,2))
plot(Connectivity[restConnectivity],meanExpr, col=as.character(colorh1),
main="Mean(Expression) vs K",xlab="Connectivity")
plot (Connectivity[restConnectivity],varExpr, col= as.character(colorh1), main="Var(Expression)
vs K" ,xlab="Connectivity") 
```

Here the rows are genes and the columns are samples.

Well defined modules results in characteristic band structures since the corresponding genes are highly correlated.

```{r}
par(mfrow=c(2,1), mar=c(1, 2, 4, 1))
ClusterSamples=hclust(dist(datExpr[,restConnectivity] ),method="average")
# for the first (turquoise) module we use
which.module="turquoise"
rg=colorpanel(50, low="green", mid="black", high="red")
heatmap(t(scale(datExpr[ClusterSamples$order,restConnectivity][,colorh1==which.module ])),col=rg,Colv=NA)
# for the second (blue) module we use
which.module="blue"
heatmap(t(scale(datExpr[ClusterSamples$order,restConnectivity][,colorh1==which.module ])),col=rg,Colv=NA)
```

Now we extend the color definition to all genes by coloring all non-module genes grey.

```{r}
color1=rep("grey",dim(datExpr)[[2]])
color1[restConnectivity]=as.character(colorh1)
# The function intramodularConnectivity computes the whole network connectivity kTotal,
# the within module connectivity (kWithin). kOut=kTotal-kWithin and
# and kDiff=kIn-kOut=2*kIN-kTotal
ConnectivityMeasures=intramodularConnectivity(ADJ,colors=colorh1)
names(ConnectivityMeasures)

# The following plots show the gene significance vs intramodular connectivity
colorlevels=levels(factor(colorh1))
par(mfrow=c(2,3),mar=c(5, 4, 4, 2) + 0.1)
for (i in c(1:length(colorlevels) ) ) {
whichmodule=colorlevels[[i]];restrict1=colorh1==whichmodule
verboseScatterplot(ConnectivityMeasures$kWithin[restrict1],
GeneSignificance[restrict1],col=colorh1[restrict1],main= paste("set I,",
whichmodule),ylab="Gene Significance",xlab="Intramodular k")
} 
```
# Plots to compare with GWENA analysis

```{r}
library(tidyverse)

ggplot(data.frame(grey=sum(colorh1=="grey"),
                   blue=sum(colorh1=="blue"),
                   turquoise=sum(colorh1=="turquoise"),
                   yellow=sum(colorh1=="yellow"),
                   green=sum(colorh1=="green"),
                   brown=sum(colorh1=="brown")) %>% 
          pivot_longer(c("grey","blue","turquoise","yellow","green","brown"),
                       names_to=c("module"),values_to="count")) +
  geom_bar(aes(x=reorder(module,-count),y=count), stat = 'identity') + 
  labs(y = "Number of genes", x="module") + theme_bw()

ggsave("figures/WGCNA_genes_per_module.png")
```

```{r plot_module_profile}
modules <- list(grey=numeric(),
                blue=numeric(),
                turquoise=numeric(),
                yellow=numeric(),
                green=numeric(),
                brown=numeric())

for(i in 1:length(colorh1))
{
  if(colorh1[i] == "grey")
  {
    modules$grey <- append(modules$grey,toString(i))
  }
  else if(colorh1[i] == "blue")
  {
    modules$blue <- append(modules$blue,toString(i))
  }
  else if(colorh1[i] == "turquoise")
  {
    modules$turquoise <- append(modules$turquoise,toString(i))
  }
  else if(colorh1[i] == "yellow")
  {
    modules$yellow <- append(modules$yellow,toString(i))
  }
  else if(colorh1[i] == "green")
  {
    modules$green <- append(modules$green,toString(i))
  }
  else if(colorh1[i] == "brown")
  {
    modules$brown <- append(modules$brown,toString(i))
  }
}

colnames(datExpr) <- 1:dim(datExpr)[2]

GWENA::plot_expression_profiles(datExpr, modules)
ggsave("figures/WGCNA_expr_profiles.png")
```

```{r modules_enrichment, fig.height=12}
enrichment <- GWENA::bio_enrich(modules)
GWENA::plot_enrichment(enrichment, interactive=F)
ggsave("figures/WGCNA_enrichment.png")
```

# Hard Clustering

```{r}
thresholds1= c(seq(.1,.5, by=.1), seq(.55,.95, by=.05) ) 
RdichotTable=WGCNA::pickHardThreshold(datExpr, cutVector=thresholds1)[[2]]
gc()
```

```{r}
cex1=0.7
gc()
par(mfrow=c(1,2))
plot(RdichotTable[,1], -sign(RdichotTable[,4])*RdichotTable[,3],xlab="Hard Threshold
tau",ylab="Scale Free Topology Model Fit,signed R^2", type="n")
text(RdichotTable[,1], -sign(RdichotTable[,4])*RdichotTable[,3] , labels=thresholds1,cex=cex1)

# this line corresponds to using an R^2 cut-off of h
abline(h=0.95,col="red")
plot(RdichotTable[,1], RdichotTable[,6],xlab="Hard Threshold tau",ylab="Mean Connectivity",
type="n")
text(RdichotTable[,1], RdichotTable[,6] , labels=thresholds1, cex=cex1) 
```

```{r}
corhelp=cor(datExpr[,restConnectivity],use="pairwise.complete.obs")
whichmodule="brown"
# the following data frame contains the intramodular connectivities
# corresponding to different hard thresholds
datconnectivitiesHard=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(thresholds1)))
names(datconnectivitiesHard)=paste("kWithinTau",thresholds1,sep="")
for (i in c(1:length(thresholds1)) ) {
datconnectivitiesHard[,i]=apply(abs(corhelp[colorh1==whichmodule,
colorh1==whichmodule])>=thresholds1[i],1,sum)}
SpearmanCorrelationsHard=signif(cor(GeneSignificance[ colorh1==whichmodule],
datconnectivitiesHard, method="s",use="p")) 
```


```{r}
datkTOMINHard=data.frame(matrix(666,nrow=sum(colorh1==whichmodule),ncol=length(thresholds1)))
names(datkTOMINHard)=paste("omegaWithinHard",thresholds1,sep="")
for (i in c(1:length(thresholds1)) ) {
  datconnectivitiesHard[,i]=
    apply(1-TOMdist(apply(abs(
      corhelp[colorh1==whichmodule,colorh1==whichmodule])>thresholds1[i],
      2,as.numeric)),1,sum)
}
SpearmanCorrelationskTOMHard=as.vector(
  signif(cor(GeneSignificance[colorh1==whichmodule], 
             datconnectivitiesHard, method="s",use="p")))
```


```{r}
ADJ= abs(cor(datExpr[,restConnectivity],use="p"))>0.7
ADJ <- apply(ADJ,2,as.numeric)
gc()
# This is the unweighted connectivity
k=as.vector(apply(ADJ,2,sum)) 

plot(k, Connectivity[restConnectivity],
      xlab="Unweighted Connectivity", ylab="Weighted Connectivity", 
     main=paste( "correlation = ",
                 signif(cor(k,Connectivity[restConnectivity]),2)),
     col=colorh1)

dissTOM=TOMdist(ADJ)
gc() 
```

Now we carry out hierarchical clustering with the TOM matrix. Branches of the resulting clustering tree will be used to define gene modules. 

```{r}
hierTOM = hclust(as.dist(dissTOM),method="average");
par(mfrow=c(1,1))
plot(hierTOM,labels=F)
colorh2=as.character(modulecolor2(hierTOM,h1=.75, minsize1=125))
table(colorh1,colorh2)

#Rand index to measure agreement between the clusterings
randIndex(table(colorh1,colorh2)) 
```

```{r}
par(mfrow=c(3,1), mar=c(2,2,2,1))
plot(hierTOM, main="Unweighted Network Module Tree ", labels=F, xlab="",
sub="");
hclustplot1(hierTOM, colorh2, main="Colored by Unweighted modules")
hclustplot1(hierTOM, colorh1, main="Colored by Weighted modules")
TOMplot(dissTOM , hierTOM, as.character(datSummary$color1[restConnectivity]))
gc() 
```